---
title: "Boston Celtics Player Analysis Current Season"
author: "Kaden"
date: "2024-12-25"
output: html_document
---
#The purpose of this analysis is to determine which player's need improvements for the current Boston Celtics 2024-2025 season.

#Loading all necessary packages for the project (you will need to install the package beforehand with the "install.packages" function)

```{r}
library(rvest) #For webscraping

library(dplyr) #For any data manipulation

library(ggplot2) #For data visualizations

library(tidyr) #For data cleaning

library(jsonlite) #Useful for API scraping

library(httr) #For handling HTTP requests

```

#Data Scraping
```{r}
boston_url <- "https://www.basketball-reference.com/teams/BOS/2025.html" #this URL contains stats that are from the trusted basketball-reference website.

boston_data <- read_html(boston_url) #This will read the HTML content.
```

#Extracting the table containing Boston player stats.
```{r}
boston_stats <- boston_data %>%
  html_node("#per_game_stats") %>%
  html_table(fill = TRUE)
```

#Preview the data
```{r}
head(boston_stats)
```



#Cleaning The Data
```{r}
boston_clean <- boston_stats %>%
  filter(!is.na(Player)) %>% #giving us all Players that are not NA
  filter(!grepl("Total", Player)) %>% #Remove the Team Totals Row
  select(Player, PTS, AST, TRB, `3P%`, `FG%`, `FT%`, MP,) %>% #Selecting only the stats we want to look at from the table
  drop_na() #Removes players with empty stats
```



#Preview the Clean Data 
```{r}
head(boston_clean)
```

#Calculating the Team Average
```{r}
boston_avg <- boston_clean %>%
  summarise(
    avg_PTS = mean(PTS),
    avg_AST = mean(AST),
    avg_TRB = mean(TRB),
    avg_3PPercent = mean(`3P%`),
    avg_FTPercent = mean(`FT%`),
    avg_FGPercent = mean(`FG%`),
    avg_MP = mean(MP)
  )

print(boston_avg)
```

#Indentify Underpeforming Players
```{r}
underperformed_plyrs <- boston_clean %>%
  filter(
  PTS < boston_avg$avg_PTS,
  AST < boston_avg$avg_AST,
  `FG%` < boston_avg$avg_FGPercent
)
print(underperformed_plyrs)
```

#Visualizations!!!!
```{r} 
boston_clean <- boston_clean %>%
  mutate(above_avg = PTS >= boston_avg$avg_PTS)
#this will showcase players who are above the Team's Average in total Points Vs. players who are Below Team's Average.

ggplot(boston_clean, aes(x = reorder(Player, PTS), y = PTS, fill = above_avg)) +
  geom_bar(stat = "identity") + 
  coord_flip() + 
  geom_text(aes(label = PTS), position = position_stack(0.5), color = 'white', fontface = "bold", size = 2.3) +
  labs(title = "Boston Celtics Points per Game vs. Team Average",
      x = "Player",
      y = "Points per Game") +
  scale_fill_manual(values = c("black", "forest green"),labels = c("Below Average", "Above Average")) +
  theme_minimal()
```


```{r}
ggplot(boston_clean, aes(x = MP, y = PTS)) +
  geom_point(aes(color = Player, size = 0.1)) +
  geom_text(
    data = subset(boston_clean, MP > 10), 
    aes(label = Player), 
    vjust = -0.5 , hjust = 1.2, size = 2.5, color = "black"
  ) +
  scale_x_continuous(limits = c(2,40), breaks = seq(2,40,5)) +
  scale_y_continuous(limits = c(0, 30), breaks = seq(0,30,2)) +
  labs(
    title = "Minutes Played Per Game Vs. Points Per Game",
    subtitle = "Relationship between MP and PPG for players",
    x = "Minutes Played per Game",
    y = "Point per Game"
  ) +
  theme_minimal() +
  theme(legend.key.size = unit(0.5, "cm"))
```



```{r}

# Create a dataset for underperforming players
underperforming <- boston_clean %>% 
  filter(Player %in% underperformed_plyrs$Player) %>%
  select(Player, MP, PTS, FG., AST)

# Reshape data to long format for faceting
underperforming_long <- underperforming %>%
  pivot_longer(cols = c(PTS, FG., AST), names_to = "Stat", values_to = "Value")

# Calculate team averages
team_avg <- boston_clean %>%
  summarise(avg_PTS = mean(PTS, na.rm = TRUE),
            avg_FG = mean(FG., na.rm = TRUE),
            avg_AST = mean(AST, na.rm = TRUE)) %>%
  pivot_longer(cols = everything(), names_to = "Stat", values_to = "Avg")

# Merge averages with underperforming players data
underperforming_long <- underperforming_long %>%
  left_join(team_avg, by = "Stat")

# Plot
ggplot(underperforming_long, aes(x = Player, y = Value, fill = Stat)) +
  geom_bar(stat = "identity", position = "dodge") + # Bars for players' performance
  geom_hline(aes(yintercept = Avg), color = "red", linetype = "dashed") + # Team Average
  facet_wrap(~ Stat, scales = "free_y") + # Separate facets for each category
  labs(
    title = "Underperforming Players vs Team Average",
    subtitle = "Comparison in PPG, FG%, and AST",
    x = "Player",
    y = "Value"
  ) +
  scale_fill_manual(values = c("PTS" = "blue", "FG." = "green", "AST" = "orange")) + # Colors for stats
  theme_minimal() +
  theme(legend.position = "top")

```
